---
title: "Corrected 3-Factor CFA Pipeline"
author: "Felicity Simpson"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
editor_options:
  chunk_output_type: console
---

# Overview

## Purpose

This document implements a 3-factor CFA pipeline for 9  cognitive indicators.

## Key analytic notes

- Univariate outliers are handled cell-wise. Values with absolute z-scores greater than 3 are replaced with `NA`, while participants are retained.
- Multivariate outliers are screened using a partial Mahalanobis approach that tolerates partially missing rows.
- The Mahalanobis cutoff is row-specific and depends on the number of observed indicators available for that participant.
- Remaining missing data are handled in the CFA using FIML.
- RT and cost variables are reverse-scored before the final CFA so that higher values reflect better performance.

# Setup

```{r setup, message=FALSE, warning=FALSE}
options(stringsAsFactors = FALSE)

library(lavaan)
```

# Helper functions

## Descriptive statistics helper

```{r helper-descriptives}
describe_simple <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) {
    return(data.frame(
      n = 0,
      mean = NA_real_,
      sd = NA_real_,
      median = NA_real_,
      min = NA_real_,
      max = NA_real_
    ))
  }
  data.frame(
    n = length(x),
    mean = mean(x),
    sd = sd(x),
    median = median(x),
    min = min(x),
    max = max(x)
  )
}

column_descriptives <- function(df) {
  out <- lapply(df, describe_simple)
  out <- do.call(rbind, out)
  out <- cbind(variable = names(df), out)
  rownames(out) <- NULL
  out
}
```

# Load data

## Read source file

```{r load-data}
dat <- read.csv("C:/Users/c3371138/Dropbox/optical-imaging/AF_optical_2_updated.csv")
```

## Select analytic sample and indicators

```{r select-analytic-sample}
keep_vars <- c(
  "record_id",
  "VRMDRTC.x", "VRMFRDS.x", "VRMIRTC.x",
  "SimpleChoiceLatency.x", "ComplexChoiceLatency.x",
  "IES_MTTICMD.x", "IES_MTTMTCMD.x",
  "IES_congruencyCost.x", "IES_mixCost.x"
)

semvars <- dat[dat$record_id >= 2001, keep_vars]

cat("\n==================== RAW DATA SUMMARY ====================\n")
print(summary(semvars))
```

# Missing data and preprocessing

## Missing data summary

```{r missing-data-summary}
missing_summary <- data.frame(
  variable = names(semvars)[-1],
  n_missing = sapply(semvars[-1], function(x) sum(is.na(x))),
  pct_missing = round(100 * sapply(semvars[-1], function(x) sum(is.na(x))) / nrow(semvars), 2)
)

cat("\n==================== MISSING DATA SUMMARY ====================\n")
print(missing_summary)
```

## Z-standardize indicators

```{r z-standardise}
analysis_data <- semvars
analysis_data[-1] <- lapply(analysis_data[-1], function(x) as.numeric(scale(x)))

cat("\n==================== DESCRIPTIVES (STANDARDIZED DATA) ====================\n")
print(column_descriptives(analysis_data[-1]))
```

## Correlation matrix before outlier handling

```{r correlations-before}
cor_mat_before <- cor(analysis_data[-1], use = "pairwise.complete.obs")

cat("\n==================== CORRELATION MATRIX BEFORE REMOVAL ====================\n")
print(round(cor_mat_before, 2))
```

## Histograms before outlier handling

```{r histograms-before, fig.width=10, fig.height=8}
op <- par(no.readonly = TRUE)
par(mfrow = c(3, 3), mar = c(3, 3, 2, 1))

for (v in names(analysis_data)[-1]) {
  hist(
    analysis_data[[v]],
    main = v,
    xlab = "Standardized score",
    col = "grey80",
    border = "white"
  )
}

par(op)
```

# Univariate outlier screening

## Flag univariate outliers

```{r univariate-flags}
uni_flag_matrix <- sapply(analysis_data[-1], function(x) abs(x) > 3)
uni_flag_matrix <- as.data.frame(uni_flag_matrix)

uni_summary <- data.frame(
  variable = names(analysis_data)[-1],
  n_univariate_outliers = sapply(analysis_data[-1], function(x) sum(abs(x) > 3, na.rm = TRUE)),
  pct_univariate_outliers = round(
    100 * sapply(analysis_data[-1], function(x) sum(abs(x) > 3, na.rm = TRUE)) / nrow(analysis_data),
    2
  )
)

cat("\n==================== UNIVARIATE OUTLIER SUMMARY (|z| > 3) ====================\n")
print(uni_summary)
```

## Participant-level univariate flag

```{r participant-univariate-flag}
analysis_data$any_univariate_outlier <- apply(uni_flag_matrix, 1, function(x) any(x, na.rm = TRUE))

cat(
  "\nParticipants with at least one univariate outlier: ",
  sum(analysis_data$any_univariate_outlier, na.rm = TRUE),
  " out of ", nrow(analysis_data), "\n",
  sep = ""
)
```

## Replace univariate outlier cells with `NA`

```{r replace-univariate-outliers}
analysis_data_na_uni <- analysis_data

indicator_names <- setdiff(names(analysis_data_na_uni), c("record_id", "any_univariate_outlier"))

for (v in indicator_names) {
  analysis_data_na_uni[[v]][abs(analysis_data_na_uni[[v]]) > 3] <- NA
}

n_cells_replaced <- sum(
  sapply(analysis_data[indicator_names], function(x) sum(abs(x) > 3, na.rm = TRUE))
)

cat("\nTotal univariate outlier cells replaced with NA: ", n_cells_replaced, "\n", sep = "")
```

## Missingness after univariate replacement

```{r missingness-after-univariate}
missing_after_uni <- data.frame(
  variable = indicator_names,
  n_missing_after_uni_replacement = sapply(analysis_data_na_uni[indicator_names], function(x) sum(is.na(x))),
  pct_missing_after_uni_replacement = round(
    100 * sapply(analysis_data_na_uni[indicator_names], function(x) sum(is.na(x))) / nrow(analysis_data_na_uni),
    2
  )
)

cat("\n==================== MISSINGNESS AFTER UNIVARIATE REPLACEMENT ====================\n")
print(missing_after_uni)
```

# Multivariate outlier screening

## Partial Mahalanobis function

```{r mahalanobis-function}
mahalanobis_partial <- function(df, id_col = "record_id", p_cut = 0.997, min_vars = 3) {

  X <- df[, setdiff(names(df), id_col), drop = FALSE]
  n <- nrow(X)

  md <- rep(NA_real_, n)
  df_used <- rep(NA_integer_, n)
  cutoff <- rep(NA_real_, n)
  ref_n <- rep(NA_integer_, n)
  outlier <- rep(NA, n)

  patterns <- apply(!is.na(X), 1, function(z) paste(z, collapse = "|"))
  unique_patterns <- unique(patterns)

  for (pat in unique_patterns) {
    idx <- which(patterns == pat)
    present <- which(strsplit(pat, "\\|")[[1]] == "TRUE")

    if (length(present) < min_vars) next

    X_sub <- X[, present, drop = FALSE]
    ref_idx <- complete.cases(X_sub)
    X_ref <- X_sub[ref_idx, , drop = FALSE]

    if (nrow(X_ref) <= length(present)) next

    S <- cov(X_ref)

    if (!is.matrix(S) || any(!is.finite(S)) || det(S) <= .Machine$double.eps) next

    mu <- colMeans(X_ref)
    md_vals <- mahalanobis(
      x = as.matrix(X_sub[idx, , drop = FALSE]),
      center = mu,
      cov = S
    )

    df_i <- length(present)
    cutoff_i <- qchisq(p_cut, df = df_i)

    md[idx] <- md_vals
    df_used[idx] <- df_i
    cutoff[idx] <- cutoff_i
    ref_n[idx] <- nrow(X_ref)
    outlier[idx] <- md_vals > cutoff_i
  }

  data.frame(
    record_id = df[[id_col]],
    mahalanobis_d = md,
    mahalanobis_df = df_used,
    mahalanobis_cutoff = cutoff,
    mahalanobis_ref_n = ref_n,
    multivariate_outlier = outlier
  )
}
```

## Apply partial Mahalanobis screening

```{r apply-mahalanobis}
md_input <- analysis_data_na_uni[, setdiff(names(analysis_data_na_uni), "any_univariate_outlier"), drop = FALSE]

md_results <- mahalanobis_partial(
  df = md_input,
  id_col = "record_id",
  p_cut = 0.997,
  min_vars = 3
)

cat("\n==================== MAHALANOBIS RESULTS ====================\n")
print(md_results)

cat("\n==================== MULTIVARIATE OUTLIER COUNTS ====================\n")
print(table(md_results$multivariate_outlier, useNA = "ifany"))
```

## Note on the bug fix

The corrected function uses a `|` separator when storing missingness patterns and splits on that same separator when identifying observed variables. This avoids the earlier pattern-matching bug where concatenated `TRUE`/`FALSE` values were split character by character rather than token by token.

## Flagged multivariate outliers

```{r flagged-outlier-ids}
outlier_ids <- md_results$record_id[which(md_results$multivariate_outlier %in% TRUE)]

cat("\nFlagged multivariate outlier IDs:\n")
print(outlier_ids)

cat("\nN with Mahalanobis distance computed:", sum(!is.na(md_results$mahalanobis_d)), "\n")
```

# Remove multivariate outliers

## Merge flags and remove flagged participants

```{r remove-multivariate-outliers}
analysis_data_cleanflags <- merge(
  analysis_data_na_uni,
  md_results[, c("record_id", "multivariate_outlier")],
  by = "record_id",
  all.x = TRUE,
  sort = FALSE
)

analysis_data_cleanflags$multivariate_outlier[is.na(analysis_data_cleanflags$multivariate_outlier)] <- FALSE
analysis_data_clean <- analysis_data_cleanflags[!analysis_data_cleanflags$multivariate_outlier, ]

cat("\n==================== SAMPLE SIZE AFTER REMOVAL ====================\n")
cat("Original N:", nrow(analysis_data_cleanflags), "\n")
cat("Removed multivariate outliers:", length(outlier_ids), "\n")
cat("Retained N:", nrow(analysis_data_clean), "\n")
```

## Descriptives after multivariate outlier removal

```{r descriptives-after-removal}
clean_indicator_names <- setdiff(
  names(analysis_data_clean),
  c("record_id", "any_univariate_outlier", "multivariate_outlier")
)

cat("\n==================== DESCRIPTIVES AFTER REMOVAL ====================\n")
print(column_descriptives(analysis_data_clean[clean_indicator_names]))
```

## Correlation matrix after multivariate outlier removal

```{r correlations-after-removal}
cor_mat_after <- cor(analysis_data_clean[clean_indicator_names], use = "pairwise.complete.obs")

cat("\n==================== CORRELATION MATRIX AFTER REMOVAL ====================\n")
print(round(cor_mat_after, 2))
```

## Histograms after multivariate outlier removal

```{r histograms-after, fig.width=10, fig.height=8}
op <- par(no.readonly = TRUE)
par(mfrow = c(3, 3), mar = c(3, 3, 2, 1))

for (v in clean_indicator_names) {
  hist(
    analysis_data_clean[[v]],
    main = v,
    xlab = "Standardized score",
    col = "grey80",
    border = "white"
  )
}

par(op)
```

# Reverse-score indicators for the final CFA

## Higher scores indicate better performance

```{r reverse-score}
reverse_vars <- c(
  "SimpleChoiceLatency.x", "ComplexChoiceLatency.x",
  "IES_MTTICMD.x", "IES_MTTMTCMD.x",
  "IES_congruencyCost.x", "IES_mixCost.x"
)

analysis_data_clean[reverse_vars] <- lapply(analysis_data_clean[reverse_vars], function(x) x * -1)
```

# Confirm final analytic sample used for CFA

## Participants retained versus used by lavaan

```{r final-sample-check}
all_missing_rows <- apply(is.na(analysis_data_clean[clean_indicator_names]), 1, all)

cat("\nRetained after multivariate outlier removal:", nrow(analysis_data_clean), "\n")
cat("Rows empty on all indicators and therefore ignored by lavaan:", sum(all_missing_rows), "\n")
cat("Expected number used by lavaan:", nrow(analysis_data_clean) - sum(all_missing_rows), "\n")
```

# CFA model

## Model specification

```{r model-specification}
cfa3_model <- '
  Verbal_three =~ VRMDRTC.x + VRMFRDS.x + VRMIRTC.x

  CognitiveCost_three =~
    IES_MTTICMD.x +
    IES_MTTMTCMD.x +
    IES_mixCost.x +
    IES_congruencyCost.x

  ProcSpeed_three =~
    b1*SimpleChoiceLatency.x + b1*ComplexChoiceLatency.x

  Verbal_three ~~ CognitiveCost_three
  Verbal_three ~~ ProcSpeed_three
  CognitiveCost_three ~~ ProcSpeed_three

  IES_MTTICMD.x ~~ IES_MTTMTCMD.x
  IES_mixCost.x ~~ IES_congruencyCost.x
'
```

## Fit CFA model with FIML

```{r fit-cfa}
cfa_input <- analysis_data_clean[clean_indicator_names]

cfa3_fit <- cfa(
  model = cfa3_model,
  data = cfa_input,
  std.lv = TRUE,
  missing = "fiml"
)

cat("\n==================== CFA SUMMARY ====================\n")
summary(cfa3_fit, fit.measures = TRUE, standardized = TRUE, rsquare = TRUE)
```

## Fit indices

```{r fit-indices}
fit_indices <- fitMeasures(
  cfa3_fit,
  c("chisq", "df", "pvalue", "rmsea", "srmr", "cfi", "tli", "aic", "bic", "pnfi")
)

print(round(fit_indices, 3))
```

## Standardized factor loadings

```{r standardized-loadings}
std_solution <- standardizedSolution(cfa3_fit)
loadings_table <- std_solution[std_solution$op == "=~", c("lhs", "rhs", "est.std", "pvalue")]
print(loadings_table)
```

## Modification indices above 3.84

```{r modification-indices}
mod_indices <- modificationIndices(cfa3_fit)
mod_indices_high <- mod_indices[mod_indices$mi > 3.84, ]
print(mod_indices_high)
```



# Factor scores and merged dataset

## Extract factor scores

```{r factor-scores}
factor_scores <- lavPredict(cfa3_fit, type = "lv")
factor_scores_df <- data.frame(record_id = analysis_data_clean$record_id, factor_scores)

cat("\n==================== FACTOR SCORE STRUCTURE ====================\n")
str(factor_scores_df)
```

## Merge factor scores back to the original dataset

```{r merge-factor-scores}
final_dataset <- merge(dat, factor_scores_df, by = "record_id", all.x = TRUE, sort = FALSE)

cat("\n==================== MERGED DATASET SUMMARY ====================\n")
cat("Rows in original data:", nrow(dat), "\n")
cat("Rows in merged data:", nrow(final_dataset), "\n")
cat("\nNew latent score variables added:\n")
print(names(factor_scores_df)[-1])
```

## Optional save step

```{r save-output, eval=FALSE}
write.csv(
  final_dataset,
  "C:/Users/c3371138/Dropbox/optical-imaging/AF_optical_3factor_c.csv",
  row.names = FALSE
)
```

# Notes for manuscript updating

## Suggested reporting points

- Multivariate outlier screening was updated to use a partial Mahalanobis procedure that screened participants with partially observed indicator sets rather than restricting screening to complete cases only.
- Mahalanobis distances were computed for participants with at least 3 observed indicators.
- Three multivariate outliers were removed after screening.
- The final CFA was estimated with FIML on the retained sample, with completely empty rows ignored automatically by `lavaan`.
- RT and cost variables were reverse-scored before model estimation so that higher scores reflected better performance.
