---
title: "Final sample diagram for manuscript"
author: "Felicity Simpson"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8
)
```

# 1. Purpose

# 1.1 Overview

This script produces a manuscript-ready sample flow diagram for the
Mediterranean diet and cerebral arterial elasticity paper (PReFx paper).
The diagram shows how the Newcastle analytic cohort (n = 198) fed into
three separate analytic samples:

- primary regression sample (listwise deletion, n = 159)
- CFA sample (partial Mahalanobis outlier screening, n = 195 retained;
  n = 190 used by lavaan)
- mediation model (FIML, n = 198)

# 1.2 Key analytic notes

- Newcastle analytic cohort is defined as record_id >= 2001.
- Primary regression exclusions reflect listwise deletion across
  MedDiet, PReFx, age, sex, education, MVPA, and energy intake.
- CFA outlier screening proceeds in two stages: (1) univariate outlier
  cells (|z| > 3) are replaced with NA while participants are retained;
  (2) multivariate outliers are identified via a partial Mahalanobis
  procedure that accommodates partially missing rows (minimum 3
  observed indicators, chi-square cutoff at p = .003).
- Three participants were excluded as multivariate outliers. Five
  participants retained in the CFA sample had no observed data on any
  cognitive indicator and were ignored automatically by lavaan.
- The mediation models were fitted on the full Newcastle cohort using
  FIML; no listwise deletion was applied.

---

# 2. Packages

```{r packages}
library(tidyverse)
library(DiagrammeR)
library(knitr)
```

---

# 3. File paths

Update these paths if needed.

```{r file-paths}
path_primary <- "C:/Users/c3371138/Dropbox/optical-imaging/AF_optical_3factor_c.csv"
path_cfa     <- "C:/Users/c3371138/Dropbox/optical-imaging/AF_optical_2_updated.csv"
path_mediate <- "C:/Users/c3371138/Dropbox/optical-imaging/AF_optical_3factor_c.csv"
```

---

# 4. Variables used in each analysis

```{r variable-lists}
primary_vars <- c(
  "record_id", "age", "sex", "anu2_total_yrs_edu", "totalmvpa", "kJwithDF",
  "MedDiet.Panagiotakis", "prefx_pfc_avg"
)

cfa_vars <- c(
  "record_id",
  "VRMDRTC.x", "VRMFRDS.x", "VRMIRTC.x",
  "SimpleChoiceLatency.x", "ComplexChoiceLatency.x",
  "IES_MTTICMD.x", "IES_MTTMTCMD.x",
  "IES_congruencyCost.x", "IES_mixCost.x"
)

mediate_vars <- c(
  "record_id", "MedDiet.Panagiotakis", "age", "sex", "anu2_total_yrs_edu",
  "totalmvpa", "kJwithDF", "prefx_pfc_avg",
  "Verbal_three", "CognitiveCost_three", "ProcSpeed_three"
)
```

---

# 5. Helper functions

```{r helpers}
# 5.1 Partial Mahalanobis outlier screening function.
# Accommodates partially missing rows by computing Mahalanobis distances
# using only the observed indicators for each participant.
# Participants with fewer than min_vars observed indicators are skipped (NA returned).
# A row-specific chi-square cutoff is applied based on the number of
# observed indicators for that participant.
mahalanobis_partial <- function(df, id_col = "record_id", p_cut = 0.997, min_vars = 3) {
  X <- df[, setdiff(names(df), id_col), drop = FALSE]
  n <- nrow(X)

  md       <- rep(NA_real_, n)
  df_used  <- rep(NA_integer_, n)
  cutoff   <- rep(NA_real_, n)
  ref_n    <- rep(NA_integer_, n)
  outlier  <- rep(NA, n)

  patterns        <- apply(!is.na(X), 1, function(z) paste(z, collapse = "|"))
  unique_patterns <- unique(patterns)

  for (pat in unique_patterns) {
    idx     <- which(patterns == pat)
    present <- which(strsplit(pat, "\\|", fixed = FALSE)[[1]] == "TRUE")

    if (length(present) < min_vars) next

    X_sub   <- X[, present, drop = FALSE]
    ref_idx <- complete.cases(X_sub)
    X_ref   <- X_sub[ref_idx, , drop = FALSE]

    if (nrow(X_ref) <= length(present)) next

    mu      <- colMeans(X_ref)
    S       <- stats::cov(X_ref)

    bad_cov <- any(is.na(S)) || isTRUE(det(S) == 0)
    if (bad_cov) next

    md_vals  <- mahalanobis(X[idx, present, drop = FALSE], center = mu, cov = S)
    cutoff_i <- qchisq(p_cut, df = length(present))

    md[idx]      <- md_vals
    df_used[idx] <- length(present)
    cutoff[idx]  <- cutoff_i
    ref_n[idx]   <- nrow(X_ref)
    outlier[idx] <- md_vals > cutoff_i
  }

  tibble(
    record_id            = df[[id_col]],
    mahalanobis_d        = md,
    df_used              = df_used,
    cutoff               = cutoff,
    ref_n                = ref_n,
    multivariate_outlier = outlier
  )
}

# 5.2 Formatting helper for node labels.
fmt_n <- function(x) format(x, big.mark = ",", scientific = FALSE)
```

---

# 6. Load data

```{r load-data}
dat_primary_raw <- read.csv(path_primary)
dat_cfa_raw     <- read.csv(path_cfa)
dat_med_raw     <- read.csv(path_mediate)
```

---

# 7. Derive sample counts

## 7.1 Primary regression branch

```{r primary-branch}
primary_newcastle <- dat_primary_raw %>%
  filter(record_id >= 2001) %>%
  dplyr::select(all_of(primary_vars))

n_primary_newcastle <- nrow(primary_newcastle)
n_primary_complete  <- sum(complete.cases(primary_newcastle))
n_primary_excluded  <- n_primary_newcastle - n_primary_complete

primary_missing_breakdown <- tibble(
  variable  = names(primary_newcastle),
  n_missing = sapply(primary_newcastle, function(x) sum(is.na(x)))
)

kable(
  primary_missing_breakdown,
  caption = "Missingness for variables used in the primary regression models"
)
```

## 7.2 CFA branch

```{r cfa-branch}
semvars <- dat_cfa_raw %>%
  filter(record_id >= 2001) %>%
  dplyr::select(all_of(cfa_vars))

n_cfa_newcastle <- nrow(semvars)

# 7.2.1 Z-standardise indicators.
analysis_data        <- semvars
analysis_data[-1]    <- lapply(analysis_data[-1], function(x) as.numeric(scale(x)))

# 7.2.2 Univariate outlier screening: flag cells with |z| > 3.
uni_flag_matrix                     <- sapply(analysis_data[-1], function(x) abs(x) > 3)
analysis_data$any_univariate_outlier <- apply(uni_flag_matrix, 1, function(x) any(x, na.rm = TRUE))

n_participants_with_uni <- sum(analysis_data$any_univariate_outlier, na.rm = TRUE)
n_uni_cells             <- sum(uni_flag_matrix, na.rm = TRUE)

# 7.2.3 Replace univariate outlier cells with NA; participants are retained.
analysis_data_na_uni <- analysis_data
indicator_names      <- setdiff(names(analysis_data_na_uni), c("record_id", "any_univariate_outlier"))

for (v in indicator_names) {
  analysis_data_na_uni[[v]][abs(analysis_data_na_uni[[v]]) > 3] <- NA
}

# 7.2.4 Partial Mahalanobis multivariate outlier screening.
md_input   <- analysis_data_na_uni[, setdiff(names(analysis_data_na_uni), "any_univariate_outlier"), drop = FALSE]
md_results <- mahalanobis_partial(df = md_input, id_col = "record_id", p_cut = 0.997, min_vars = 3)

outlier_ids         <- md_results$record_id[which(md_results$multivariate_outlier %in% TRUE)]
n_md_computed       <- sum(!is.na(md_results$mahalanobis_d))
n_multivar_outliers <- length(outlier_ids)

# 7.2.5 Remove multivariate outliers.
analysis_data_cleanflags <- merge(
  analysis_data_na_uni,
  md_results[, c("record_id", "multivariate_outlier")],
  by      = "record_id",
  all.x   = TRUE,
  sort    = FALSE
)

analysis_data_cleanflags$multivariate_outlier[
  is.na(analysis_data_cleanflags$multivariate_outlier)
] <- FALSE

analysis_data_clean <- analysis_data_cleanflags[
  !analysis_data_cleanflags$multivariate_outlier,
]

n_cfa_retained <- nrow(analysis_data_clean)

# 7.2.6 Identify rows that are empty on all indicators and will be ignored by lavaan.
clean_indicator_names     <- setdiff(
  names(analysis_data_clean),
  c("record_id", "any_univariate_outlier", "multivariate_outlier")
)
all_missing_rows          <- apply(is.na(analysis_data_clean[clean_indicator_names]), 1, all)
n_all_missing_after_retention <- sum(all_missing_rows)
n_cfa_used_by_lavaan      <- n_cfa_retained - n_all_missing_after_retention
```

```{r cfa-branch-summary}
cfa_summary <- tibble(
  step = c(
    "Newcastle analytic cohort for CFA",
    "Participants with at least one univariate outlier cell",
    "Univariate outlier cells replaced with NA",
    "Rows with Mahalanobis distance computed",
    "Multivariate outliers removed",
    "Retained after multivariate outlier removal",
    "Rows empty on all indicators (ignored by lavaan)",
    "Observations used by lavaan"
  ),
  n = c(
    n_cfa_newcastle,
    n_participants_with_uni,
    n_uni_cells,
    n_md_computed,
    n_multivar_outliers,
    n_cfa_retained,
    n_all_missing_after_retention,
    n_cfa_used_by_lavaan
  )
)

kable(cfa_summary, caption = "CFA sample derivation summary")

if (length(outlier_ids) > 0) {
  cat("Flagged multivariate outlier record IDs:", paste(outlier_ids, collapse = ", "), "\n")
}
```

## 7.3 Mediation branch

```{r mediation-branch}
med_newcastle <- dat_med_raw %>%
  filter(record_id >= 2001) %>%
  dplyr::select(all_of(mediate_vars))

n_med_newcastle <- nrow(med_newcastle)

med_missing_breakdown <- tibble(
  variable  = names(med_newcastle),
  n_missing = sapply(med_newcastle, function(x) sum(is.na(x)))
)

kable(
  med_missing_breakdown,
  caption = "Missingness for variables entering mediation models fitted with FIML"
)
```

---

# 8. Key counts summary

```{r key-counts}
key_counts <- tibble(
  measure = c(
    "Newcastle analytic cohort",
    "Primary regression sample (listwise deletion)",
    "Excluded from primary regression (missing data)",
    "CFA: retained after outlier screening",
    "CFA: excluded as multivariate outliers",
    "CFA: rows empty on all indicators (ignored by lavaan)",
    "CFA: observations used by lavaan",
    "Mediation model (FIML)"
  ),
  n = c(
    n_primary_newcastle,
    n_primary_complete,
    n_primary_excluded,
    n_cfa_retained,
    n_multivar_outliers,
    n_all_missing_after_retention,
    n_cfa_used_by_lavaan,
    n_med_newcastle
  )
)

kable(key_counts, caption = "Key sample counts for the manuscript flow diagram")
```

---

# 9. Manuscript flow diagram

Two versions are provided below. Version A orders branches from most to
least restrictive (Primary | CFA | Mediation). Version B reverses this
order (Mediation | CFA | Primary). Run both and select the layout that
reads most clearly in your target journal format.

Exclusion boxes use dashed borders to follow CONSORT/STROBE convention:
inclusions flow straight down the main spine; exclusions branch
sideways.


The following caption can be adapted for the manuscript. It carries the
methodological detail that has been kept out of the diagram boxes.

**Figure 1.** *Flow diagram for the final analytic samples.*
The Newcastle analytic cohort comprised 198 participants (record_id
>= 2001). For the primary regression analyses, the analytic sample was
derived by listwise deletion across Mediterranean diet score, PReFx,
age, sex, years of education, MVPA, and total energy intake, yielding
n = 159 complete cases. For the confirmatory factor analysis (CFA),
cognitive indicators were first screened for univariate outliers
cell-wise (|z| > 3; 16 cells replaced with NA across 14 participants),
followed by multivariate outlier screening using a partial Mahalanobis
procedure that accommodated partially observed rows (minimum 3 observed
indicators; chi-square cutoff at p = .003). Three participants were
excluded as multivariate outliers, leaving n = 195 retained. Of these,
five had no observed data on any cognitive indicator and were ignored
automatically by lavaan, yielding n = 190 observations used for model
estimation. The mediation models were estimated on the full Newcastle
cohort (n = 198) using Full Information Maximum Likelihood (FIML),
with no listwise deletion applied.



```{r diagram-version-A, results='asis'}

label_newcastle    <- paste0("Newcastle analytic cohort\nn = ", fmt_n(n_primary_newcastle))

label_primary      <- paste0("Primary analysis\n(listwise deletion)\nn = ", fmt_n(n_primary_complete))
label_primary_excl <- paste0("Excluded: missing data\nn = ", fmt_n(n_primary_excluded))

label_cfa_retained <- paste0("Retained after outlier screening\nn = ", fmt_n(n_cfa_retained))
label_cfa_excl     <- paste0("Excluded as multivariate outliers\nn = ", fmt_n(n_multivar_outliers))
label_cfa_lavaan   <- paste0("CFA: observations used by lavaan\nn = ", fmt_n(n_cfa_used_by_lavaan))
label_cfa_empty    <- paste0("Empty on all cognitive indicators\nn = ", fmt_n(n_all_missing_after_retention))
label_mediation    <- paste0("Mediation model fitted with FIML\nn = ", fmt_n(n_med_newcastle))

DiagrammeR::grViz(glue::glue("
  digraph sample_flow_A {{
    graph [layout = dot, rankdir = TB, nodesep = 1.5, ranksep = 0.9]
    node [shape = box, style = rounded, fontname = Helvetica, fontsize = 11, margin = 0.25, width = 2.8]
    edge [fontname = Helvetica, fontsize = 10]

    A  [label = '{label_newcastle}']

    B1 [label = '{label_primary}']
    B2 [label = '{label_primary_excl}', style = 'rounded,dashed']

    C1 [label = '{label_cfa_retained}']
    C2 [label = '{label_cfa_excl}',  style = 'rounded,dashed']
    C3 [label = '{label_cfa_lavaan}']
    C4 [label = '{label_cfa_empty}', style = 'rounded,dashed']
    C5 [label = '{label_mediation}']

    A  -> B1
    A  -> C1

    B1 -> B2 [style = dashed]

    C1 -> C2 [style = dashed]
    C1 -> C3
    C3 -> C4 [style = dashed]
    C3 -> C5

    {{rank = same; B1; C1}}
    {{rank = same; B2; C2}}
  }}
"))
```




