---
title: "CFA-3-factors"
author: "Felicity"
date: "2024-11-25"
output: html_document
---

  
```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(lavaan)
library(psych)
library(datawizard)
library(performance)
dat <- read.csv("C:/Users/c3371138/Dropbox/optical-imaging/dat/AF_optical_2.csv")
```

#The following variables will be considered for analysis:  VRMDRTC, VRMFRDS, VRMIRTC, SimpleChoiceLatency, ComplexChoiceLatency, IES_MTTICMD, IES_MTTMTCMD, IES_congruencyCost, IES_mixCost 

```{r}
semvars <- dat %>% filter(record_id >= 2001) %>% 
  select(record_id,
         # Verbal Memory
         VRMDRTC, VRMFRDS, VRMIRTC, 
         # Processing Speed
         SimpleChoiceLatency, ComplexChoiceLatency,
         # Cognitive Cost (executive function in manuscript)
         IES_MTTICMD, IES_MTTMTCMD, 
         IES_congruencyCost, IES_mixCost
         #crystallised intelligence
        # educationtotal, NIH_Reading, NIH_PictureVocab
  )

summary(semvars)
```

# Three-factor CFA based on theoretical model from Kałamała et al. (2025)
# https://www.biorxiv.org/content/10.1101/2025.04.06.646611v1
# Factors: Verbal Memory, Executive Function, Processing Speed
# Equality constraint on Processing Speed loadings per Crawford & Lamarre (2021)


```{r}
library(ggplot2)
library(tidyr)
library(dplyr)

# 1. Reshape data from wide to long
semvars_long <- semvars %>%
  select(-record_id) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  )

# 2. Calculate mean and standard error for each variable
stats_before <- semvars_long %>%
  group_by(variable) %>%
  summarize(
    mean = mean(value, na.rm = TRUE),
    se   = sd(value, na.rm = TRUE)/sqrt(n()),
    .groups = "drop"
  )

# 3. Plot each variable in its own facet with free x-scales
ggplot(stats_before, aes(y = 0, x = mean)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = mean - se, xmax = mean + se), height = 0.2) +
  facet_wrap(~ variable, scales = "free_x", ncol = 3) +
  labs(
    title = "Mean ± SE (Before Scaling)",
    x = "Mean Value",
    y = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.y  = element_blank(),   # Hide the y-axis labels (since y=0 is just a placeholder)
    axis.ticks.y = element_blank()
  )

```
```{r}
# Histogram for each variable with free scales for clarity
ggplot(semvars_long, aes(x = value)) +
  geom_histogram(bins = 30, fill = "grey", color = "black") +
  facet_wrap(~ variable, scales = "free_x") +
  labs(
    title = "Histograms (Before Scaling)",
    x = "Value",
    y = "Frequency"
  ) +
  theme_minimal()

```




```{r}
# Scale each variable individually, preserving NAs
scaled_data <- semvars %>%
  mutate(across(where(is.numeric) & !all_of("record_id"), scale))
```

# Winsorize data
Winsorize at ±3.29 SD (p < .001 in normal distribution)
Conservative threshold to remove only extreme outliers 

```{r}
# Winsorize non-missing values only (preserving NAs)
winsorized <- scaled_data %>%
  select(-record_id) %>%
  winsorize(method = "zscore", robust = TRUE, threshold = 3.29)

# Reattach record_id
winsorized <- cbind(record_id = scaled_data$record_id, winsorized)
```

Check for univariate outliers. 

```{r}
# Check univariate outliers (this step does not change the data)
uni_outliers <- check_outliers(winsorized %>% select(-record_id), method = "zscore_robust")
print(uni_outliers)  # Expect: OK or very few
```
No extreme univariate outliers remain. 

#Multivariate outliers check

```{r}
# Detect multivariate outliers on complete rows only
complete_cases <- winsorized %>% drop_na()
mult_outliers <- check_outliers(complete_cases %>% select(-record_id), method = "mcd")
outlier_rows <- attributes(mult_outliers)$outlier_count$all$Row
outlier_ids <- complete_cases$record_id[outlier_rows]
```

```{r}
mult_outliers
```
Let's remove the multivariate outliers. 

```{r}
# Remove multivariate outliers by record_id
cleaned_data <- winsorized %>% filter(!record_id %in% outlier_ids)

# Reverse-score RT and cost variables to align directionality
cleaned_data <- cleaned_data %>%
  mutate(across(c(SimpleChoiceLatency, ComplexChoiceLatency,
                  IES_MTTICMD, IES_MTTMTCMD,
                  IES_congruencyCost, IES_mixCost),
                ~ . * -1))
```



```{r}
# Remove record_id and pivot longer for visualization
cleaned_data_long <- cleaned_data %>% 
  select(-record_id) %>% 
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

# Calculate summary statistics per variable (ignoring NAs)
stats_after <- cleaned_data_long %>% 
  group_by(variable) %>% 
  summarize(
    mean = mean(value, na.rm = TRUE),
    se = sd(value, na.rm = TRUE)/sqrt(sum(!is.na(value))),
    .groups = "drop"
  )

# Error bar plot (mean ± SE) after scaling and reverse scoring
ggplot(stats_after, aes(x = variable, y = mean)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  labs(title = "Mean plus/minus SE (After Scaling and Reverse Scoring)",
       x = "Cognitive Variable",
       y = "Scaled Mean Value") +
  theme_minimal(base_size = 14) +
  coord_flip()



```
```{r}
# Histogram for each variable after scaling and reverse scoring
ggplot(cleaned_data_long, aes(x = value)) +
  geom_histogram(bins = 30, fill = "grey", color = "black") +
  facet_wrap(~ variable, scales = "free") +
  labs(title = "Histograms (After Scaling and Reverse Scoring)",
       x = "Scaled Value",
       y = "Frequency") +
  theme_minimal(base_size = 14)


```
#Winsorisation summary
```{r}


# Remove record_id for comparison
scaled_matrix <- as.matrix(scaled_data[ , !names(scaled_data) %in% "record_id"])
winsorized_matrix <- as.matrix(winsorized[ , !names(winsorized) %in% "record_id"])

# Count number of values that were winsorised
num_winsorised <- sum(scaled_matrix != winsorized_matrix, na.rm = TRUE)

# Count total number of data points
total_values <- sum(!is.na(scaled_matrix))

# Percentage of values winsorised
perc_winsorised <- num_winsorised / total_values * 100

cat("Number of winsorised values:", num_winsorised, "\n")
cat("Total data points:", total_values, "\n")
cat("Percentage winsorised:", round(perc_winsorised, 2), "%\n")

```

### Retained FIML CFA model used for manuscript and factor extraction
### Full model

```{r}
cfa3 <- '
Verbal_three =~ VRMDRTC + VRMFRDS + VRMIRTC
CognitiveCost_three =~ IES_MTTICMD + IES_MTTMTCMD + IES_mixCost + IES_congruencyCost
ProcSpeed_three =~ b1*SimpleChoiceLatency + b1*ComplexChoiceLatency

# Correlations
Verbal_three ~~ CognitiveCost_three
Verbal_three ~~ ProcSpeed_three
CognitiveCost_three ~~ ProcSpeed_three

IES_MTTICMD ~~ IES_MTTMTCMD 
IES_mixCost ~~ IES_congruencyCost

'
```

# Note: Little's MCAR test conducted in separate script showed p = .380
# Supporting use of FIML for missing data handling


```{r}
cfa3fit <- cfa(cfa3, data = cleaned_data, std.lv = TRUE, missing = 'fiml')
summary(cfa3fit)

```
Participants missing every variable are dropped
cfa3 results for the three-factor model after removing the residual covariance between IES_mixCost and IES_congruencyCost. 


```{r}
# List of fit indices
indices <- c("chisq", "df", "pvalue", "rmsea", "srmr", "cfi", "pnfi")

# Get the requested fit measures
fitMeasures(cfa3fit, indices)
```
Model fit is good and seems better than a four-factor model with general intelligence
For reference these were my OLD fit indices
#  chisq     df pvalue  rmsea   srmr    cfi   pnfi 
## 67.192 47.000  0.028  0.051  0.047  0.966  0.639






```{r}
# Retrieve modification indices without specifying a threshold
mod_indices <- modificationIndices(cfa3fit)

# View modification indices with a cut-off value for the modification index (e.g., > 3.84)
mod_indices_high <- mod_indices[mod_indices$mi > 3.84, ]

# Print the filtered modification indices
```


```{r}
print(mod_indices_high)



```









```{r}
# Extract factor scores for latent variables
factor_scores <- lavPredict(cfa3fit, type = "lv")

# Bind the factor scores with the original data to create a new dataframe
new_dataframe <- cbind(cleaned_data, factor_scores)

# View the structure of the new dataframe to confirm the addition
str(new_dataframe)

# Merge the new data with the original dataset
final_dataset <- left_join(dat, new_dataframe, by = "record_id")

# Save the updated dataset to a new CSV to avoid overwriting the original accidentally
#write.csv(final_dataset, "C:/Users/c3371138/Dropbox/optical-imaging/AF_optical_3factor.csv", row.names = FALSE)

# Print confirmation message
print("The updated dataset with new latent variables has been successfully saved.")

```
```{r}

### Alternative possible solution?

# - uses the winsorized data before multivariate outlier removal, 
# keeps the usual direction alignment so that higher scores reflect better performance, and - # retains only the MTT within-task residual  covariance.

winsorized_reversed <- winsorized %>%
  mutate(across(c(SimpleChoiceLatency, ComplexChoiceLatency,
                  IES_MTTICMD, IES_MTTMTCMD,
                  IES_congruencyCost, IES_mixCost),
                ~ . * -1))

cfa_fiml_alternative <- '
Verbal_three =~ VRMDRTC + VRMFRDS + VRMIRTC
CognitiveCost_three =~ IES_MTTICMD + IES_MTTMTCMD + IES_mixCost + IES_congruencyCost
ProcSpeed_three =~ b1*SimpleChoiceLatency + b1*ComplexChoiceLatency

# Latent factor correlations
Verbal_three ~~ CognitiveCost_three
Verbal_three ~~ ProcSpeed_three
CognitiveCost_three ~~ ProcSpeed_three

# Within-task residual covariance for MTT only
IES_MTTICMD ~~ IES_MTTMTCMD
'

fit_fiml_alternative <- cfa(
  cfa_fiml_alternative,
  data = winsorized_reversed,
  std.lv = TRUE,
  missing = "fiml"
)

cat("\n====================================================\n")
cat("Alternative FIML model explored during model checking\n")
cat("Potential alternative solution; NOT used for factor extraction\n")
cat("====================================================\n\n")

summary(
  fit_fiml_alternative,
  fit.measures = TRUE,
  standardized = TRUE,
  ci = TRUE
)
```




#------------------------------------------------------------------------------


```{r}
### old version -->  Alternative FIML model without TSWT residual covariance (Heywood case)
#What was in the previous version of the manuscript (sent July 2025)

cfa_heywood <- '
Verbal_three =~ VRMDRTC + VRMFRDS + VRMIRTC
CognitiveCost_three =~ IES_MTTICMD + IES_MTTMTCMD + IES_mixCost + IES_congruencyCost
ProcSpeed_three =~ b1*SimpleChoiceLatency + b1*ComplexChoiceLatency

# Correlations
Verbal_three ~~ CognitiveCost_three
Verbal_three ~~ ProcSpeed_three
CognitiveCost_three ~~ ProcSpeed_three

IES_MTTICMD ~~ IES_MTTMTCMD
'

fit_heywood <- cfa(
  cfa_heywood,
  data = cleaned_data,
  std.lv = TRUE,
  missing = "fiml"
)

cat("\n====================================================\n")
cat("Comparison model 2: Alternative FIML model without TSWT residual covariance (Heywood case)\n")
cat("NOT used for factor extraction\n")
cat("====================================================\n\n")

summary(
  fit_heywood,
  fit.measures = TRUE,
  standardized = TRUE,
  ci = TRUE
)
```


```{r}
# Question 1: raw correlations between the EF measures

ef_corr <- cleaned_data %>%
  select(IES_MTTICMD, IES_MTTMTCMD, IES_congruencyCost, IES_mixCost) %>%
  cor(use = "pairwise.complete.obs")

round(ef_corr, 3)
```

```{r}
# Clean long-format table for Word

library(tidyr)
library(dplyr)
library(psych)

ef_data <- cleaned_data %>%
  select(IES_MTTICMD, IES_MTTMTCMD, IES_congruencyCost, IES_mixCost)

ef_corr_test <- psych::corr.test(ef_data, use = "pairwise")

corr_table <- as.data.frame(ef_corr_test$r) %>%
  tibble::rownames_to_column("Measure_1") %>%
  pivot_longer(-Measure_1, names_to = "Measure_2", values_to = "r") %>%
  mutate(
    p = as.data.frame(ef_corr_test$p) %>%
      tibble::rownames_to_column("Measure_1") %>%
      pivot_longer(-Measure_1, names_to = "Measure_2", values_to = "p") %>%
      pull(p)
  ) %>%
  filter(Measure_1 != Measure_2)

corr_table %>%
  mutate(
    r = round(r, 3),
    p = round(p, 3)
  )
```

